Two Problems for Sophistication

### 2015-01-21 Facticity/Facticity/paper.tex

For the first argument, we take a generative perspective. We can generate data from a model in $i \in \K$ by feeding it random bits until it produces an argument, we will call the resulting probability distribution for model $p_i(x) = \sum_{y:\phi_i(y) = x} 2^{-|y|}$. In this scenario, a sophistication could be called \emph{consistent} if, for data of sufficiently high complexity, it accurately reflects the complexity of that distribution from which it originated.\footnotemark Suppose that the data issampled from  universal distribution: let $\phi_u(\bar\imath y)=\phi_i(y)$ and sample from $p_u$ as described above. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of the resulting data?

\footnotetext{While in general we do not wish to make any assumptions about the origin of the data, in some cases, it can make sense to do so as a sanity check.}

@@

The second reason we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Figure~\ref{fig:diagram} provides an illustration. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in the figure. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. \emph{However}, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can always occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that \emph{part} of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=-\log p_i(x)-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data---if $\phi_i$ truly generated the data, then the probability that the randomness deficiency is more than $k$ bits is less than $2^{-k}$ \cite{bloem2014safe}[Lemma~6]. In the Monet example above, this will allow us to disqualify the model expressing that the data is actually, say, a recording of Jazz music.

>>>>

For the first argument, we take a generative perspective. We can generate data from a model in $\phi_i, i \in \K$ by feeding it random bits until it produces an output. We will call the resulting probability distribution $p_i(x) = \sum_{y:\phi_i(y) = x} 2^{-|y|}$. In this scenario, a sophistication could be called \emph{consistent} if, for data of sufficiently high complexity, it accurately reflects the complexity of that distribution from which it originated. Suppose that the data issampled from  universal distribution: let $\phi_u(\bar\imath y)=\phi_i(y)$ and sample from $p_u$ as described above. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of the resulting data?

@@

The second reason we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in Figure~\ref{fig:diagram}. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. However, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can always occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that part of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=-\log p_i(x)-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data---if $p_i$ truly generated the data, then the probability that the randomness deficiency is more than $k$ bits is less than $2^{-k}$ \cite{bloem2014safe}[Lemma~6]. In the Monet example above, this will allow us to disqualify the model expressing that the data is actually, say, a recording of Jazz music.
